Causes and effects in Dichotomous Comparative Judgments: an information-theoretical system with plausible mechanism

### 2024-07-25 _tex/index.tex
Built site for gh-pages

\begin{abstract}
Dichotomous Comparative Judgment (DCJ) requires judges to compare pairs
of stimuli to determine which one exhibits a higher degree of a specific
trait. DCJ has proven effective and reliable across various fields
\citet{Pollitt_2012b}; \citet{Jones_2015}; \citet{vanDaal_et_al_2016};
\citet{Bartholomew_et_al_2018}; \citet{Lesterhuis_2018};
\citet{Bartholomew_et_al_2020}; \citet{Marshall_et_al_2020};
\citet{Boonen_et_al_2020}{]}. However, despite the method's widespread
use, existing literature lacks a clear explanation of the complexities
and assumptions underpinning the DCJ system, as well as the plausible
mechanisms through which DCJ data are generated. This study addresses
these issues by integrating DCJ into the framework of causal inference.
Specifically, utilizing a structural approach to causal inference, the
study develops a scientific model to clarify the causal assumptions and
mechanisms inherent in the DCJ system. It then translates this model
into a probabilistic statistical framework to estimate statistical
relationships and infer causal connections within the system. This
research builds upon Thurstone's law of comparative judgment
\citeyearpar{Thurstone_1927} and provide a robust probabilistic
foundation for the statistical analysis of DCJ data. The findings of
this study offer valuable insights for researchers and analysts involved
in education and assessment procedures who implement or design DCJ
experiments.
\end{abstract}

@@

DCJ has proven effective in assessing competencies and traits
predominantly within the educational realm, as demonstrated by
\citet{Pollitt_2012b}, \citet{Jones_2015}, \citet{vanDaal_et_al_2016},
\citet{Bartholomew_et_al_2018}, \citet{Lesterhuis_2018},
\citet{Bartholomew_et_al_2020}, and \citet{Marshall_et_al_2020}.
However, its application transcends education, as exemplified by
\citet{Boonen_et_al_2020}. The methodology has also evolved to include
multiple, as opposed to pairwise comparisons
\citep{Luce_1959, Placket_1975}, and to accommodate comparisons with
ordinal outcomes \citep{Tutz_1986, Agresti_1992}. Overall, research
suggests that DCJ offers an alternative and efficient approach to
measurement and evaluation, characterized by its reliability and
validity \citep{Lesterhuis_2018, vanDaal_2020, Marshall_et_al_2020}.
Nevertheless, despite the method's widespread use, the literature does
not offer a clear representation of the plausible mechanisms that
generate DCJ data. Particularly, there is no depiction of the complexity
and the underlying assumptions of the DCJ system, nor how different
assessment factors can potentially influence the observed DCJ outcome.

According to \citet{Verhavert_et_al_2019} and \citet{vanDaal_2020},
several assessment factors interact and influence the method's outcome.
These factors include the number and characteristics of the stimuli,
their \emph{proximity} in terms of the assessed trait, the number of
comparison per stimulus, and the pairing algorithm used. Furthermore,
since the method relies on judges' assessments, the number and
characteristics of judges, their \emph{discrimination} abilities, and
the number of comparisons per judge also play pivotal roles. Moreover,
when the stimuli represent sub-units of higher-levels units, factors
such as the number and characteristics of these units, along with their
\emph{proximity} in terms of the assessed trait, can significantly
influence the outcome. For example, \citet{vanDaal_et_al_2016} assessed
the academic writing skills of university students (units) using
multiple argumentative essays (sub-units).

Although several studies have examined the individual impact of these
factors on the method's reliability
\citep{Bramley_2015, Pollitt_2012b, Bramley_et_al_2019, Verhavert_et_al_2019, Crompvoets_et_al_2022, vanDaal_et_al_2017, Gijsen_et_al_2021},
none, to the best of the authors' knowledge, have provided a transparent
depiction of the DCJ system and the mechanisms generating the DCJ
outcome. This study aims to fill this gap by representing DCJ within the
causal inference framework. Specifically, using the structural approach
to causal inference \citep{Wright_1921, Pearl_2009, Pearl_et_al_2016},
the study aims to construct a scientific model. This model will
elucidate the underlying assumptions of the DCJ system, providing
plausible mechanisms for how the DCJ outcome is generated. Next, using a
minimal set of assumptions, the study will translate the scientific
model into a probabilistic statistical model. This model will produce
statistical estimates to draw inferences about plausible causal
relationships within the DCJ system.

Ultimately, this research builds upon Thurstone's law of comparative
judgment \citeyearpar{Thurstone_1927} and provide a robust probabilistic
foundation for the statistical analysis of DCJ data. Consequently, the
findings of this study offer valuable insights for researchers and
analysts involved in education and assessment procedures who implement
or design DCJ experiments.

@@

However, the structural approach offers a key advantage over the
potential outcomes approach: it enables the graphical representation of
systems through directed acyclic graphs
\citep[DAG,][]{Gross_et_al_2018, Neal_2020}. In this context, DAGs
fulfill the role of a heuristic, effectively conveying the assumed
causal structure of a system. They do not represent detailed statistical
models but allow researchers to deduce which statistical models can
provide valid causal inferences, assuming the causal structure depicted
in the DAG is accurate \citep{McElreath_2020}.

>>>>

\begin{abstract}
Dichotomous Comparative Judgment (DCJ) requires judges to compare pairs
of stimuli to determine which one exhibits a higher degree of a specific
trait. DCJ has proven effective and reliable across various fields
\citep{Pollitt_2012b, Jones_2015, vanDaal_et_al_2016, Bartholomew_et_al_2018, Lesterhuis_2018, Bartholomew_et_al_2020, Marshall_et_al_2020, Boonen_et_al_2020}.
However, despite the method's widespread use, existing literature lacks
a clear explanation of the complexities and assumptions underpinning the
DCJ system, as well as the plausible mechanisms through which DCJ data
are generated. This study addresses these issues by representing DCJ
within the framework of causal inference. Specifically, utilizing a
structural approach to causal inference, the study develops a scientific
model to clarify the causal assumptions and mechanisms inherent in the
DCJ system. It then translates this model into a probabilistic
statistical framework to estimate statistical relationships and infer
causal connections within the system. This research provides a robust
probabilistic foundation for the statistical analysis of DCJ data,
building upon Thurstone's law of comparative judgment
\citeyearpar{Thurstone_1927}. Its findings offer valuable insights for
researchers and analysts designing and implementing DCJ experiments.
\end{abstract}

@@

DCJ has proven effective in assessing competencies and traits
predominantly within the educational realm, as demonstrated by
\citet{Pollitt_2012b}, \citet{Jones_2015}, \citet{vanDaal_et_al_2016},
\citet{Bartholomew_et_al_2018}, \citet{Lesterhuis_2018},
\citet{Bartholomew_et_al_2020}, and \citet{Marshall_et_al_2020}.
However, its application transcends education, as exemplified by
\citet{Boonen_et_al_2020}. The methodology has also evolved to include
multiple, as opposed to pairwise comparisons
\citep{Luce_1959, Placket_1975}, and to accommodate comparisons with
ordinal outcomes \citep{Tutz_1986, Agresti_1992}. Overall, research
suggests that DCJ offers an alternative and efficient approach to
measurement and evaluation, characterized by its reliability and
validity \citep{Lesterhuis_2018, vanDaal_2020, Marshall_et_al_2020}.
Nevertheless, despite the method's widespread use, existing literature
lacks a clear representation of the plausible mechanisms through which
DCJ data are generated. Particularly, there is no depiction of the
complexity and the assumptions underpinning the DCJ system, nor how
different assessment factors can potentially influence the observed DCJ
outcome.

According to \citet{Verhavert_et_al_2019} and \citet{vanDaal_2020},
several assessment factors interact and influence the method's outcome.
These factors include the number and characteristics of the stimuli,
their \emph{proximity} in terms of the assessed trait, the number of
comparison per stimulus, and the pairing algorithm used. Furthermore,
since the method relies on judges' assessments, the number and
characteristics of judges, their \emph{discrimination} abilities, and
the number of comparisons per judge also play pivotal roles. Moreover,
when the stimuli represent sub-units of higher-levels units, factors
such as the number and characteristics of these units, along with their
\emph{proximity} in terms of the assessed trait, can significantly
influence the outcome. For instance, \citet{vanDaal_et_al_2016} assessed
academic writing skills of university students (units) using multiple
argumentative essays (sub-units).

Although several studies have examined the individual impact of these
factors on the method's reliability
\citep{Bramley_2015, Pollitt_2012b, Bramley_et_al_2019, Verhavert_et_al_2019, Crompvoets_et_al_2022, vanDaal_et_al_2017, Gijsen_et_al_2021},
none, to the best of the authors' knowledge, have provided a transparent
depiction of the DCJ system and the mechanisms generating the DCJ
outcome. This study aims to fill this gap by representing DCJ within the
framework of causal inference. Specifically, utilizing a structural
approach to causal inference
\citep{Wright_1921, Pearl_2009, Pearl_et_al_2016}, the study develops a
scientific model to clarify the causal assumptions and mechanisms
inherent in the DCJ system. Next, using a minimal set of assumptions,
the study translates the scientific model into a probabilistic
statistical model. This model aims to produce statistical estimates to
draw inferences about plausible causal relationships within the DCJ
system.

Ultimately, this research provides a robust probabilistic foundation for
the statistical analysis of DCJ data, building upon Thurstone's law of
comparative judgment \citeyearpar{Thurstone_1927}. Consequently, its
findings offer valuable insights for researchers and analysts designing
and implementing DCJ experiments.

@@

However, the structural approach offers a key advantage over the
potential outcomes approach: it enables the graphical representation of
systems through directed acyclic graphs
\citep[DAG,][]{Gross_et_al_2018, Neal_2020}. DAGs are heuristics that
effectively convey the assumed causal structure of a system. They do not
represent detailed statistical models but allow researchers to deduce
which statistical models can provide valid causal inferences, assuming
the causal structure depicted in the DAG is accurate
\citep{McElreath_2020}.
